---
title: "PA1_template.Rmd"
date: "03/07/2020"
output: md_document
---
## Loading and preprocessing the data
```{r setup, echo=TRUE,message=F}
library(plyr)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(knitr)
unzip("activity.zip")
activity<-read.csv("activity.csv",header = T)
activity$date<-as.Date(as.character(activity$date),"%Y-%m-%d")
activity<-mutate(activity,weekday = weekdays(date))
head(activity)
str(activity)
```

## What is mean total number of steps taken per day?

Here I have used dplyr package for grouping and summarizing steps datewise.Then I have taken only those rows which does not have NA and finally I have calculated mean and median using mean() and median() function.

```{r, echo=TRUE}
activity_sum_mean_median<-activity%>%group_by(date)%>%summarise(date_wise_sum=sum(steps,na.rm = T),date_wise_mean=mean(steps,na.rm = T),date_wise_median=median(steps,na.rm = T))
activity_tibble_nona<-activity_sum_mean_median[complete.cases(activity_sum_mean_median),]
mean(activity_tibble_nona$date_wise_sum,na.rm = T)
median(activity_tibble_nona$date_wise_sum,na.rm = T)
```

## What is the average daily activity pattern?

The average daily pattern can be visualised by analysing several plots:-

1.Histogram for date wise total steps(with NA):-

```{r pressure, echo=T}
hist(activity_tibble_nona$date_wise_sum,xlab="Date wise sum of steps",ylab = "Frequency of steps",ylim = c(0,40),labels = T,main = "Histogram for date wise total steps(with NA)")
```

2.Date wise mean steps for each day with the maximum average steps mentioned on the plot
```{r, echo=T}
with(activity_tibble_nona,plot(date,date_wise_mean,type="o",pch=19,cex=1.3,xlab="Date",ylab = "Mean of steps per day"))
max_mean<-which.max(activity_tibble_nona$date_wise_mean)
text(x=activity_tibble_nona$date[max_mean],y=activity_tibble_nona$date_wise_mean[max_mean],labels =round(activity_tibble_nona$date_wise_mean[max_mean],2),cex = 0.8,adj = -0.2)
title("Date wise mean steps for each day")
```

3.A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r,echo=T}
activity1<-activity%>%group_by(interval)%>%summarize(interval_wise_mean=mean(steps,na.rm = T),interval_wise_sum=sum(steps,na.rm = T))
activity1
plot(activity1$interval,activity1$interval_wise_mean,type="l",pch=19,cex=0.8,xlab="5-min interval",ylab = "Mean of steps per day")
title("Time series plot for interval wise mean steps")
```

## Imputing missing values

Calculatation of total number of missing values in the dataset (i.e. the total number of rows with NA|NAs):-
```{r,echo=TRUE}
colSums(is.na(activity))         #total NA values in each column  
mean(is.na(activity$steps))*100  #percentage of NA values from overall dataset
```

First taking all indexes where the NA values exists in the date_wise_mean column in activity_sum_mean_median dataframe.
```{r,echo=T}
na_indexes<-which(is.na(activity_sum_mean_median$date_wise_mean))
na_indexes
```

## Devise a strategy for filling in all of the missing values in the dataset.
The stategy that I have used here is as follows:-
 In the date_wise_mean column in activity_sum_mean_median dataframe,the NA values are assigned the mean of the values that precedes and succeeds that particular NA value.For example,suppose NA value in date_wise_mean column is at 8th row so this NA takes the value of mean of the value at 7th row and value at 9th row. If NA value is at 1st row then mean of values at next two successor rows is taken and if the last observation is NA then mean of values at two row that precedes the last row.
 
 Suppose 1st row is NA then mean of value at 2nd and 3rd row will be taken and suppose if last row(61 in our case) is NA then mean of value at 60th and 59th row is taken as subsitute for that NA value.
 
 Suppose that if two NA values occure simoultaneously (in our case at 40th and 41st row),then value at 40 is the mean of value at places 29th row and 38th row, and the value at 41st place will be the mean of values at 42nd and 43rd place.

```{r,echo=TRUE,message=F}
for(i in 1:nrow(activity_sum_mean_median)){
  if(is.na(activity_sum_mean_median$date_wise_mean[i]) && (i==1||i==41)){
    activity_sum_mean_median$date_wise_mean[i]= mean(c(activity_sum_mean_median$date_wise_mean[i+1],activity_sum_mean_median$date_wise_mean[i+2]))
  }
  else if(is.na(activity_sum_mean_median$date_wise_mean[i]) && (i==61||i==40)){
    activity_sum_mean_median$date_wise_mean[i]= mean(c(activity_sum_mean_median$date_wise_mean[i-1],activity_sum_mean_median$date_wise_mean[i-2]))
  }
  else if(is.na(activity_sum_mean_median$date_wise_mean[i]))
    activity_sum_mean_median$date_wise_mean[i]= mean(c(activity_sum_mean_median$date_wise_mean[i-1],activity_sum_mean_median$date_wise_mean[i+1]))
  else 
    next
  
  }
activity_sum_mean_median
```
In the above data frame all the NA values are taken care of and we will use this mean to generate random numbers in our original dataset to fill the NA values and then we will use floor() function to convert it to an integer.

As we can see in our riginal dataset ,almost at all dates first 20-25 rows and last 20-25 rows are having 0 steps so we will do the same to substitute our NA values.Wherever NA value starts the first 21 rows from that ith row will take value 0 and last 21 rows will also take value 0 and rest 245 rows will take floor of the random values with mean as the corrosponding value is the date_wise_mean column in activity_sum_mean_median dataframe.

```{r,echo=T,message=F,warning=F}
set.seed(123)
activity3<-activity
j=1
i=1
while(i<=nrow(activity3)){
      if(is.na(activity3$steps[i])){
        activity3$steps[i:(i+20)]=0
        activity3$steps[(i+21):(i+266)]=floor(rnorm(245,mean = activity_sum_mean_median$date_wise_mean[na_indexes[j]],sd=4))
        activity3$steps[(i+267):(i+287)]=0
        j=j+1
        
           if(j>length(na_indexes))
               break
        i=i+288
      }
    else{
      i=i+1
      }
}
head(activity3,30)          #1st 21 rows are replaced by 0 as intended
activity3[78:100,]          #let's see any random row to see if were successful at what we wanted
colSums(is.na(activity3))   #we can see all nas has been replaced
```

## Are there differences in activity patterns between weekdays and weekends?

Let's create a new dataframe with one extra column for weekday or weekend and plot panels for analysing behaviour on weekdays and weekend. 
```{r,echo=T}
activity4<-mutate(activity3,weekday_or_weekend=ifelse(weekday=="Saturday","weekend",ifelse(weekday=="Sunday","weekend","weekday")))
head(activity4)

activity4_total<-activity4%>%group_by(interval,weekday_or_weekend)%>%summarize(interval_wise_mean=mean(steps))
activity4_total

g<-ggplot(activity4_total,aes(interval,interval_wise_mean))+geom_line(color="red")+facet_grid(weekday_or_weekend~.)+labs(x="Interval",y="Mean value in intervals")+ggtitle("Analysing differences in activity patterns between weekdays and weekends")
g
```
 
From the plot at weekends the interval of 1000-2000 has a rise as compared to weekdays, and for 0-1000 interval steps taken on weekdays exceeds weekends.















